---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

[![Build Status](https://travis-ci.org/aclemen1/modulr.svg)](https://travis-ci.org/aclemen1/modulr)

# modulr — A Dependency Injection (DI) Framework for R

*Until the release of the forthcoming version 0.2.0, the documentation of 
the package is a work in progress. Thank you for your understanding.*

## Description

The `modulr` package is a Dependency Injection (DI) Framework for R. 
By design, `modulr` allows to break down sequential programs into discrete, 
modular units that are loosely coupled, simple to develop, test, reuse and 
share in a wide range of situations. As every DI framework, it aims for a clear 
separation between code complication and complexity, highlighting the 
core purpose and behaviour of objects (application code), and hiding 
their construction and wiring (infrastructure code). 

## Advantages

  * modules are easy (and fun) to develop,
  * modules are easy to debug,
  * modules are easy to test,
  * modules are easy to read,
  * modules are easy to reuse,
  * modules are easy to share,
  * modules are easy to maintain, and
  * modules force (a bit) to keep up with good practices.

`modulr` is deeply inspired from [AngularJS](https://angularjs.org/) and 
[RequireJS](http://requirejs.org) for Javascript, 
as well as [guice](https://github.com/google/guice) for Java.

## Installation

<!---
You can install:

* the latest released version from CRAN with

``` r
install.packages("modulr")
```
* the latest released version from Github with
``` r
if (packageVersion("devtools") < 1.8) {
  install.packages("devtools")
}
devtools::install_github("aclemen1/modulr")
```

* the latest _bleeding edge_ development version from Github with
-->
``` r
if (packageVersion("devtools") < 1.8) {
  install.packages("devtools")
}
devtools::install_github("aclemen1/modulr@devel")
```

If you encounter a clear bug, please [file a minimal reproducible example](https://github.com/aclemen1/modulr/issues).

## A short example

To get started with `modulr`, let us consider the following situation. 
Suppose that a university needs to compute its student-teacher ratio. 
This requires to gather at least a dataset about students and a 
dataset about teachers. Due to the organization of the university, 
suppose furthermore that these datasets are accessible, 
kept and/or maintained by different people. Alice, say, 
knows everything about students, when teachers have no secret for Bob. 
To start with our calculation of a student-teacher ratio, let's ask Alice to 
provide us with a usable dataset.

```{r}
library(modulr)

# This module provides a dataset relating students and their inscriptions to courses.
# Alice is the maintainer of this module.
"data/students" %provides%
  function() {
    students <- data.frame(
      id = c(1, 2, 2, 3, 3, 3),
      course = c("maths", "maths", "physics", "maths", "physics", "chemistry"),
      stringsAsFactors = F)
    return(students)
  }
```

The anatomy of this module is very simple: "data/student" is its name and the 
body of the function following the `%provides%` operator (which is part of a 
_syntactic sugar_ for the more verbose function `define`) contains its core 
functionality, namely returning the required data frame.

It is important to note that no intrinsic computation took place in this 
**definition** process. The DI framework simply **registered** the module, 
relaying the actual evaluation of its body to another **instanciation** stage, 
as we'll see below.

In parallel, let's ask Bob to provide us with a similar module.

```{r}
# This module provides a dataset relating teachers and their courses.
# Bob is the maintainer of this module.
"data/teachers" %provides%
  function() {
    teachers <- data.frame(
      id = c(1, 2, 3),
      course = c("maths", "physics", "chemistry"),
      stringsAsFactors = F)
    return(teachers)
  }
```

Now that we have these two modules at our disposal, let's combine them into 
another module that returns a (bad) student-teacher ratio.

```{r}
"bad_stat/student_teacher_ratio" %requires%
  list(
    students = "data/students",
    teachers = "data/teachers"
  ) %provides%
  function(students, teachers) {
    ratio <- length(unique(students$id)) / length(unique(teachers$id))
    return(ratio)
  }
```

The `%requires%` operator allows us to specify the modules we rely on for the 
calculation we provide. This list of **dependencies** assigns some arbitrary and
ephemeral names to the required modules. These are those names that are then 
used to call objects into which the results of the required modules are 
**injected**, and available for use in the body of the module's definition.

It is now time to see the DI framework in action.

```{r}
ratio %<=% "bad_stat/student_teacher_ratio"
```

We say that the `%<=%` operator **instanciates** the module given on its 
right-hand side. Obviously, there are three modules involved in this process, 
namely `[data/student]` and `[data/teachers]` which are independent on a first 
_layer_, and `[bad_stat/student_teacher_ration]` which depends on them on a 
second layer. Under the hood, the framework figures out the directed acyclic 
graph (DAG) of the dependencies and computes a topological sort, grouped by 
independent modules into layers. All the modules are then evaluated in order and
the final result is assigned to the left-hand side of the `%<=%` operator.

```{r}
print(ratio)
```

## Code of Conduct

This project adheres to the [Open Code of Conduct](http://todogroup.org/opencodeofconduct/#modulr/alain.clement-pavon@unil.ch). By participating, you are expected to honor this code.
