--- 
title: 
  "Introduction to modulr" 
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: > 
  %\VignetteIndexEntry{Introduction to modulr} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}{inputenc} 
---

```{r, echo = FALSE, message = FALSE, results = 'hide'} 
set.seed(1234L) 
modules_path <- file.path(
  if (grepl("/docs/", getwd(), fixed = TRUE)) file.path("..", "..") else "..", 
  "inst", "modules")
library(modulr) 
library(networkD3)
library(chorddiag)
library(RColorBrewer)
library(memoise)
library(devtools)
options(knitr.duplicate.label = 'allow')
`%<=%` <- modulr::`%<=%` 
Sys.setlocale("LC_TIME", "en_DK.UTF-8") 
Sys.setenv(TZ = 'UTC') 
knitr::opts_chunk$set(
  collapse = TRUE, 
  comment = "#>", 
  fig.path = "./figures/modulr-",
  fig.width = 6.0, 
  fig.height = 4.0, 
  out.width = "90%", 
  fig.align = "center"
) 
BUILD <- 
  identical(tolower(Sys.getenv("BUILD")), "true") &&
  !identical(tolower(Sys.getenv("TRAVIS")), "true") && 
  identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(purl = BUILD)
gears_path <- file.path(tempdir(), "gears")
unlink(gears_path, recursive = TRUE)
options(modulr.gears_path = gears_path)
reset <- function() {
  modulr::reset()
  root_config$set(modules_path)
}
```

# Anatomy of a module

Modules are defined in a declarative way, using the keywords `%requires%`
and `%provides%`, and have four main components:

* A __name__, which identifies the module. 
* A list of required __dependencies__, if any. 
* A __provider__, which expresses the module's feature(s). 
* Optional __docstrings__, intended to document the internals of a module.

A typical module looks like the following:

```r
"name_of_module" %requires% list(

  # A list of dependencies.
  dependency_1 = "name_of_dependency_1", 
  dependency_2 = "name_of_dependency_1",
  ... = "..."

) %provides% {

  #' A recommended docstring intended to document the internals of the module.

  # A section where add-on packages are loaded and attached.
  library(package_1)
  library(package_2)
  library(...)

  # Some code that uses the objects `dependency_1` and `dependency_2" 
  # returned by the modules "name_of_dependency_1" and "name_of_dependency_2".
  object <- { ... }

  # A resulting object, which can be directly consumed or in turn injected as a 
  # dependency.
  return(object)

} 
```

When a module is defined, modulr has to _make_ it in order to evaluate the code
it provides:

```r
result <- make("name_of_module") 
```

or with a handy _syntactic sugar_:

```r
result %<=% "name_of_module" 
```

or interactively with the `hit` function:

```r
hit(name_of_module) # or
hit(name_of) # which will prompt the user to choose among all possible match
```

The result contains the computed object exposed by the module. Under the
hood, the dependencies have been sorted and appropriately made, and their 
resulting objects injected where required.

# A first working example

```{r, echo=FALSE, results='hide', message=FALSE} 
reset() 
```

Let us start by defining some modules and dependencies:

```{r} 
"foo" %provides% "Hello"

"bar" %provides% "World"

"baz" %provides% "!"

"foobar" %requires% list(
  f = "foo", 
  b = "bar",
  z = "baz"
) %provides% { 
  #' Return a concatenated string. 
  paste0(f, ", ", tolower(b), z) 
} 
```

Use `info` to output the docstrings:

```{r} 
info("foobar") 
```

Use `lsmod` to list all defined modules and their properties in a data frame:

```{r} 
lsmod(cols = c("name", "type", "dependencies", "uses", "size", "modified")) 
```

In this example, `foobar` relies on three dependencies, and `foo`, `bar` and `baz` are both injected once. 
Use `plot_dependencies` to see these relations:

```{r plot-foobar-deps} 
plot_dependencies() 
```

```{r plot-foobar-deps-chord} 
plot_dependencies(render_engine = chord_engine) 
```

Use `make` to get the resulting object provided by the module:

```{r} 
make("foobar")
```

Voilà! All the depencencies have been evaluated, injected and processed to return the expected `"Hello, world!"`. As a matter of fact, the dependencies form a [directed acyclic graph](https://en.wikipedia.org/wiki/Directed_acyclic_graph) which is topologically sorted, just in time to determine a well ordering for their evaluation before injection.

```{r} 
lsmod(cols = c("name", "type", "dependencies", "uses", "size", "modified")) 
```

# Types of modules

## Singletons

```{r, echo=FALSE, results='hide', message=FALSE}
reset()
```

All modules are singletons: once evaluated, they always return the same resulting object. This is one of the great advantages of modulr: module evaluation takes place parsimoniously, when changes are detected or explicitely required, à la [GNU Make](https://www.gnu.org/software/make).

```{r}
"timestamp" %provides% {
  #' Return a string containing a timestamp.
  format(Sys.time(), "%H:%M:%OS6")
}
```

Successive `make` calls on the module will not imply its re-evaluation:

```{r}
make("timestamp")
with_verbosity(0, make("timestamp")) # temporarily change the verbosity of make
```

Notice the `with_verbosity` wrapper around the call. To force re-evaluation, just `touch` the module:

```{r}
touch("timestamp")
make("timestamp")
```

Any change of the module's definition (even its docstrings) will be detected:

```{r}
"timestamp" %provides% {
  #' Return a string containing a timestamp with more information.
  format(Sys.time(), "%Y-%m-%d %H:%M:%OS6")
}

make("timestamp")
```

## Prototypes

```{r, echo=FALSE, results='hide', message=FALSE}
reset()
```

It is granted that all modules are singletons. Nonetheless, a module is allowed to return any object, in particular it can return a function (closure) that itself returns a desired object or produces some side effect. In this case, such a module behaves like a so-called _prototype_.

```{r}
"timestamp" %provides% {
  function() format(Sys.time(), "%H:%M:%OS6")
}
```

```{r}
make("timestamp")()

with_verbosity(0L, make("timestamp")())
```

It is important to emphasize that the module is still a singleton: the second `make` call doesn't re-evaluate. But the function that is returned by the module is itself re-evaluated each time it is called. 

## Memoised prototypes

```{r, echo=FALSE, results='hide', message=FALSE}
reset()
```

Singletons produce cached objects at make-time and prototypes produce computed objects at run-time. In a complementary manner, memoised modules produce cached objects at run-time. [Memoisation](https://en.wikipedia.org/wiki/Memoization) and [Hadley Wickam's memoise package](https://github.com/hadley/memoise) give an elegant solution to this requirement.

To see the essence of what is happening, we decrease the verbosity of modulr and set up a simple starting scenario: `foo` requires the somewhat resource-consuming `timestamp` module, defined as a singleton:

```{r}
set_verbosity(1L) # messages are shown only when changes occur

"timestamp" %provides% {
  # This is a singleton.
  message("'timestamp' is evaluated after a (short) pause...")
  Sys.sleep(1L)
  format(Sys.time(), "%H:%M:%OS6")
}

"foo" %requires% list(
  timestamp = "timestamp"
) %provides% {
  "foo"
}

system.time(make("foo"))
```

In this example, `timestamp` is evaluated even though it is not explicitely used by `foo`. It just computes a timestamp after a short pause, but it could be virtually very resource-consuming at make-time.

Let us re-define `timestamp` as a prototype:

```{r}
"timestamp" %provides% {
  # This is a prototype.
  function() {
    message("'timestamp' is evaluated after a (short) pause...")
    Sys.sleep(1L)
    format(Sys.time(), "%H:%M:%OS6")
  }
}

system.time(make("foo"))
```

Here, the evaluation consists of defining a function that pauses for a while and returns a timestamp, only when the function is explicitely called. Even if the computation encapsulated by the function is very resource-consuming, no evaluation of the returned function takes place at make-time.

Finally, let us re-define `timestamp` as a memoised module:

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(memoise)
```

```{r}
"timestamp" %provides% {
  # This is a memoised module.
  memoise::memoise(
    function() {
      message("'timestamp' is evaluated after a (short) pause...")
      Sys.sleep(1L)
      format(Sys.time(), "%H:%M:%OS6")
    }
  )
}

system.time(make("foo"))
```

The `timestamp` module returns a function which will be evaluated only when explicitely called at run-time. Let us re-define `foo` in order that it effectively uses `timestamp`.

```{r}
"foo" %requires% list(
  timestamp = "timestamp"
) %provides% {
  message("It is ", timestamp())
  "foo"
}

system.time(make("foo"))
```

Here, a timestamped message is outputed. Let us force the re-evaluation of `foo`.

```{r}
touch("foo")
system.time(make("foo"))
```

The memoised version of `timestamp` is evaluated only at run-time, not at make-time; moreover, the string containing the actual timestamp is computed only once and then cached for future calls, avoiding re-evaluation.

To force re-evaluation of the memoised function exposed by `timestamp`, use `memoise::forget`.

```{r}
memoise::forget(make("timestamp"))
touch("foo")
system.time(make("foo"))
```

## Lists

It is often useful for a module to expose several (immutable, cf. infra) objects at once by returning a list.

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
reset()
```

```{r}
"timestamps" %provides% {
  now <- function() Sys.time()
  list(
    origin = structure(0L, class = "Date"),
    yesterday = function() now() - 86400L,
    now = now,
    tomorrow = function() now() + 86400L
  )
}

ts %<=% "timestamps"

ts$origin
ts$yesterday()
ts$now()
ts$tomorrow()
```

## Environments

It is often useful for a module to expose several mutable objects at once by returning an environment.

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
reset()
```

```{r}
"configuration" %provides% {
  env <- new.env(parent = emptyenv())
  env$shape <- "circle"
  env$color <- "blue"
  env$size <- 13L
  env
}

config %<=% "configuration"

config$color
config$color <- "red"
config$color
```

This kind of module can be used to share mutable data between modules, without polluting the Global Environment.

```{r}
"widget_A" %requires% list(
  config = "configuration"
) %provides% {
  list(
    switch_color = function()
      config$color <- if (config$color == "blue") "red" else "blue"
  )
}

"widget_B" %requires% list(
  config = "configuration"
) %provides% {
  list(
    switch_shape = function()
      config$shape <- if (config$shape == "circle") "square" else "circle"
  )
}

widget_A %<=% "widget_A"
widget_B %<=% "widget_B"

widget_A$switch_color()
config$color

widget_B$switch_shape()
config$shape
```

The modulr package implements the dedicated syntactic sugar `%provides_options%` for this frequent purpose.

```{r}
undefine("configuration")

"configuration" %provides_options% list(
  shape = "circle",
  color = "blue",
  size = 13L
)

config %<=% "configuration"
widget_A %<=% "widget_A"

config$color
widget_A$switch_color()
config$color
```

It is also possible to use the __shared environment__ associated to every injector:

```{r}
"widget_B_prime" %provides% {
  list(
    switch_shape = function()
      .SharedEnv$shape <- 
        if (.SharedEnv$shape == "circle") "square" else "circle"
  )
}

widget_B_prime <- make()

.SharedEnv$shape <- "circle"
widget_B_prime$switch_shape()
.SharedEnv$shape
```


## Byte-compiled modules

Since a module is allowed to expose any R object, there is no restriction on byte-compiled code.

```{r}
"hello" %provides% {
  compiler::compile("Hello, world!")
}

eval(make())
```

```{r}
"lapply_old" %provides% {
  # Old R version of lapply.
  function(X, FUN, ...) {
    FUN <- match.fun(FUN)
    if (!is.list(X))
      X <- as.list(X)
    rval <- vector("list", length(X))
    for(i in seq(along = X))
      rval[i] <- list(FUN(X[[i]], ...))
    names(rval) <- names(X) # keep `names' !
    return(rval)
  }
}

"lapply_old/compiled" %requires% list(
  lapply_old = "lapply_old"
) %provides% {
  compiler::cmpfun(lapply_old)
}

lapply_old %<=% "lapply_old"
lapply_old_compiled %<=% "lapply_old/compiled"

system.time(for (i in 1L:10000L) lapply_old(1L:10L, is.null))
system.time(for (i in 1L:10000L) lapply_old_compiled(1L:10L, is.null))
```

# Semantic Versioning (SemVer)

The modulr package offers [Semantic Versioning](http://semver.org) capabilities: 
every module can live in several versions numbers of the form `x.y.z`, where _x_, _y_, 
and _z_ are the __major__, __minor__, and __patch__ versions, respectively. 
For instance, `foo#1.2.3` designates module _foo_ in version _1.2.3_.

Given a version number, increment the:

* __major__ version when you make incompatible changes,
* __minor__ version when you refactor and/or add functionality in a backwards-compatible manner, and
* __patch__ version when you make backwards-compatible bug fixes.

For instance, `foo#1.2.3` becomes `foo#1.2.4` after a bug fix and `foo#1.3.0` after a functionality bump.

Use:

* `~x.y.z` to refer to the most up-to-date available patch version above _x.y.z_ and __allow bug fixes, but nothing else__,
* `^x.y.z` or `^x.y` to refer to the most up-to-date available minor version above _x.y_ and __allow bug fixes and new functionalites, but nothing else__, and
* `>=x.y.z`, `>=x.y`, or `>=x` to refer to the most up-to-date version above _x.y.z_, _x.y_, or _x_, and __live on the edge of developpments__.

Here are some examples among `foo#1.2.3`, `foo#1.2.4`, `foo#1.3.0`:

* `foo#~1.2.0` refers to `foo#1.2.4`,
* `foo#~1.2.5` refers to nothing,
* `foo#^1.2.5` and `foo#^1.1` refer to `foo#1.3.0`,
* `foo#^1.3.1` and `foo#^1.4` refer to nothing, and
* `foo#>=1.1.0`, `foo#>=1.5`, and `foo#>=0` (aka _latest_) refer to `foo#1.3.0`.

## Initial scenario: no versioning

There is a good chance that your initial scenario contains no versioned module.

```{r}
"great_module" %provides% {
  function() {
    Sys.sleep(1L)
    "great features"
  }
}
```

```{r}
"complex_module" %requires% list(
  great = "great_module"
) %provides% {
  function() cat(paste("complex module using", great()))
}
```

```{r}
system.time(make("complex_module")())
```

In this scenario, `great_module` does what it is supposed to do, but clearly not very efficiently. 
You then decide to work on a new version that improves its performance.

## Setting-up versioning

First, we clone `great_module` with an initial version number.

```{r}
"great_module#0.1.0" %clones% "great_module"
```

We then adapt the requirements where `great_module` is injected as a dependency: for `complex_module`, we decide to accept bug fixes, refactorisations, and new functionalities, as long as the API does not change in an incompatible backward manner.

```{r}
"complex_module" %requires% list(
  great = "great_module#^0.1.0"
) %provides% {
  function() cat(paste("complex module using", great()))
}
```

```{r}
system.time(make("complex_module")())
```

Here is the minor bump of `great_module`, which is a little bit more efficient.

```{r}
"great_module#0.2.0" %provides% {
  # Improved internals, same interface
  function() "great optimisd features"
}
```
```{r}
system.time(make("complex_module")())
```

And here is the latest bug fix correcting the typo.

```{r}
"great_module#0.2.1" %provides% {
  # Bug fix
  function() "great optimised features"
}
```

```{r}
make("complex_module")()
```


# Location of modules

Modules can be defined in several locations: in-memory, on-disk in its own file or along another module's file, and remotely on GitHub's Gist or via the HTTP(S) protocol.

## In-memory

This is the most direct method to define a module. This is also the most volatile, 
since the lifespan of the module is limited to the R session.

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
reset()
```

```{r}
"foo" %provides% "bar"
lsmod(cols = c("name", "storage", "along", "filepath", "url"))
```

## On-disk

```{r include=FALSE}
reset()
```

This is the way to go when a module is intended to be reused. In such a case, 
the definition takes place in a dedicated R, R Markdown, or R Sweave file, which 
path and name are closely related to the module's name.

For instance, the following module definition is stored in the R file `swissknife.R`, under the sub-directory `vendor/tool` of the `./modules` directory.

```{r, code=readLines(file.path(modules_path, "vendor", "tool", "swissknife.R")), eval = FALSE}
```

When a module is invoked, modulr searches for it in-memory first, then on-disk if necessary. There are several default __root__ places where modulr looks for the module's file: `./modules/`, `./module/`, `./libs/`, `./lib/`, and `./`. This behaviour can be configured with the help of `root_config`.

```{r}
root_config$get_all()
```

This explains why modulr finds the module `vendor/tool/swissknife` under the file `./modules/vendor/tool/swissknife.R`.

```{r}
my_swissknife %<=% "vendor/tool/swissknife"
```

This also works with R Markdown `.Rmd` and R Sweave `.Rnw` files.

```{r echo=FALSE, comment='', highlight=TRUE}
cat(readLines(file.path(modules_path, "vendor", "tool", "multitool.Rmd")), sep = "\n")
```

```{r}
load_module("vendor/tool/multitool") # load only, do not make
```

```{r}
lsmod(cols = c("name", "storage", "along", "filepath", "url"))
```

Along a principal module, it is possible to define other related modules, for instance mock-ups and testing modules (cf. infra).

## Remote (_modulr gears_)

```{r include=FALSE}
reset()
```

Using GitHub's [Gist](https://gist.github.com) is a simple way to share modules with others. To illustrate this, let us consider the following remote module, aka __modulr gear__: [https://gist.github.com/aclemen1/3fcc508cb40ddac6c1e3](https://gist.github.com/aclemen1/3fcc508cb40ddac6c1e3).

```{r, eval=BUILD}
"modulr/vault" %imports% "https://gist.github.com/aclemen1/3fcc508cb40ddac6c1e3"
```

Notice that only specifiying the gist ID in `"modulr/vault" %imports% "3fcc508cb40ddac6c1e3"` has the same effect. It is possible to import modules from any URL using the HTTP(S) protocol.

Once imported, a remote module appears to be in-memory defined. 

```{r}
lsmod(cols = c("name", "storage", "along", "filepath", "url"))
```

To use a remote module as a dependency, just import it where needed (even in a remote module).

```{r, eval=BUILD}
"modulr/vault" %imports% "3fcc508cb40ddac6c1e3"

"module/using/a/gear" %requires% list(
  vault = "modulr/vault"
) %provides% {
  vault$decrypt(
    secret = "TWUnCkRAlP70XvmRlnAFrw==",
    key = "EaJWzAZjjphu9CoA+MPUVCL8mmMAGp0j6Nbga29kV/A=")
}

make()
```

Notice that sharing a module is as easy as sending this one-liner code snippet:

```r
library(modulr); "modulr/vault#^0.1.0" %imports% "3fcc508cb40ddac6c1e3"
```

Finally, private Gists and [GitHub Enterprise](https://enterprise.github.com) users are also covered, thanks to GitHub's [Personal Access Tokens](https://github.com/blog/1509-personal-api-tokens) (PAT). For instance, with the GitHub Enterprise instance of the University of Lausanne:

```r
# Set 'GITHUB_PAT' in your '.Renviron' file or right here:
# Sys.setenv(GITHUB_PAT = "Your Personal Access Token here")

"modulr/private_GitHubEnterprise_module" %imports% 
  "https://github.unil.ch/api/v3/gists/1afa4770670975d70806c2153aac50a9"
```

```{r, eval=BUILD, echo=FALSE}
(function() {
  GITHUB_PAT_bak <- Sys.getenv("GITHUB_PAT")
  Sys.setenv(GITHUB_PAT = Sys.getenv("GITHUB_UNIL_PAT"))
  on.exit(Sys.setenv(GITHUB_PAT = GITHUB_PAT_bak))
  "modulr/private_GitHubEnterprise_module" %imports% 
    "https://github.unil.ch/api/v3/gists/1afa4770670975d70806c2153aac50a9"
})()
```

Let us assume that the following module is worth publishing:

```{r, results='hide'}
"modulr/release_gist_example" %provides% "Hello World!"
```

To release this module as a modulr gear on GitHub's Gist, simply use `release_gear_as_gist`:

```r
release_gear_as_gist()
```

```{r, echo=FALSE, eval=BUILD}
g <- release_gear_as_gist(browse = FALSE)
```

```{r, include=FALSE, eval=!BUILD}
g <- list(html_url = "<URL>")
```

The module is then publicly available here: `r g[["html_url"]]`.

# Special variables

## `.Last.name`

When a module is defined, touched, or made, its name is always assigned to `.Last.name`.
The special variable `.Last.name` is also used as a default parameter for `make`, `touch`, and `undefine`.

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
reset()
```

```{r}
"foo" %provides% "bar"

.Last.name

make()

touch()

undefine()
```

## Module's metadata

```{r echo=FALSE, results='hide', message=FALSE, warning=FALSE, cache=FALSE}
reset()
```

Every module has access to some of its metadata: name, version, file path (when on-disk), etc. 
The following module illustrates this feature and is self-explanatory.

```{r, code=readLines(file.path(modules_path, "my", "great", "module", "reflection.R")), eval = FALSE}
```

```{r}
with_verbosity(0L, make("my/great/module/reflection"))
```


# The special module `modulr`

The modulr package defines a special module named `modulr` that can be injected in any
module. The purpose of this special module is to give access to useful helper 
functions related to the module into which it is injected.

```{r}
info("modulr")
```

## Messages

TODO

## Post-evaluation hook

There are situations where a post-evaluation hook is needed. For instance, to define 
an ephemeral module that can be evaluated only once, or to define a so-called _no-scoped_ 
module, which looks like a pure singleton, but behaves like a prototype.

```{r}
"ephemeral" %requires% list(
  modulr = "modulr"
) %provides% {
  modulr$post_evaluation_hook(undefine("ephemeral"))
  "A butterfly"
}

make("ephemeral") # returns a butterfly
try(make("ephemeral"), silent = TRUE) # no more
cat(geterrmessage())
```

```{r}
"no_scoped" %requires% list(
  modulr = "modulr"
) %provides% {
  modulr$post_evaluation_hook(touch("no_scoped"))
  Sys.time()
}

make("no_scoped")
Sys.sleep(1L)
make("no_scoped")
```

Notice that the expression passed to the hook is evaluated in the environment in which the module is used. Therefore, a direct call to `.__name__` would not return the name of the intuitively expected module. The following example illustrates how to circumvent this kind of difficulty.

```{r}
"no_scoped" %requires% list(
  modulr = "modulr"
) %provides% {
  eval(substitute(modulr$post_evaluation_hook(touch(me)), list(me = .__name__)))
  Sys.time()
}

make("no_scoped")
Sys.sleep(1L)
make("no_scoped")
```

# Scripting

Turning a bunch of modules working perfectly well together into a script is a 
very common situation, that can be handled with the help of the following boilerplate code:

```r
# filepath: ./script.R
"script" %requires% list(
  dep_1 = "dependency_1",
  ...
) %provides% {
  function() {
    # body of the script here
  }
}

if (.__name__ == "__main__") 
  # execute only if sourced/run as a script (à la Python)
  make()()
```

# Coding style
