--- 
title: 
  "Using modulr in packages" 
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: > 
  %\VignetteIndexEntry{Using modulr in packages} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}{inputenc} 
---

```{r, echo = FALSE, message = FALSE, results = 'hide'}
set.seed(1234L)
modules_path <- file.path(
  if (grepl("/docs/", getwd(), fixed = TRUE)) file.path("..", "..") else "..",
  "inst", "modules")
# library(modulr)
library(networkD3)
library(chorddiag)
library(RColorBrewer)
library(memoise)
library(devtools)
options(knitr.duplicate.label = 'allow')
`%<=%` <- modulr::`%<=%`
Sys.setlocale("LC_TIME", "en_DK.UTF-8")
Sys.setenv(TZ = 'UTC')
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "./figures/modulr-",
  fig.width = 6.0,
  fig.height = 4.0,
  out.width = "90%",
  fig.align = "center"
)
BUILD <-
  identical(tolower(Sys.getenv("BUILD")), "true") &&
  !identical(tolower(Sys.getenv("TRAVIS")), "true") &&
  identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(purl = BUILD)
gears_path <- file.path(tempdir(), "gears")
unlink(gears_path, recursive = TRUE)
options(modulr.gears_path = gears_path)
reset <- function() {
  modulr::reset()
  modulr::root_config$set(modules_path)
}
```

A package can import modulr for its own usage. There are then essentially two different and complementary ways to exploit all of the modulr features: for the limited scope of the _internals_ of the package itself and/or for an exported usage.

# Internal scope

In this use case, modules are defined and used _inside_ the package's namespace, without interference with any usage that could be made of modulr _outside_ of the package. Something ressembling the following lines of code should then appear in your `./R` package files:

```r
# File: ./R/init.R

#' @import modulr
NULL

my_package_injector <- get_default_injector()
set_verbosity(0L)
```

Notice that we are using [Roxygen2](https://github.com/klutometis/roxygen)'s `@import` tag here to easily populate the `NAMESPACE` file of the package with an appropriate `import` field (the `@importFrom` tag could also import only a selected subset of exported methods from modulr), so that the prefix `modulr::` can be omitted in front of modulr methods (in particular for syntactic sugars like `%provides%` and `%requires%`, see _infra_).

It is also important to make sure that this portion of code is loaded before defining and using modules, which is guaranteed by the `@include` tag (or by manually specifying a collation order in the `NAMESPACE` file), as shown in the following example:

```r
# File: ./R/greetings.R

#' @include init.R
NULL

"hello" %provides% "Hello"

"world" %provides% "World"

"greetings" %requires% list(
  hello = "hello",
  world = "world"
) %provides% {
  paste0(hello, ", ", tolower(world), "!")
}

#' @export
say_hello <- function() {
  with_injector(my_package_injector, make("greetings"))
}
```

Notice the `with_injector` wrapper around `make("greetings")` in the body of the exported `say_hello` function: since `say_hello` is suscpetible to be called from outside of the package's namespace, it is important to specify explicitely that the `make` call relies on the package's default injector captured in the `./R/init.R` file. 

It is also possible to use the injector's `get` method:

```r
#' @export
say_hello <- function() {
  my_package_injector$get("greetings")
}
```

Notice that none of the modules is directly exposed by the package:

```r
library(my_package)
(modulr::lsmod())
```

```{r, echo = FALSE}
modulr::with_verbosity(0L, modulr::reset())
(modulr::lsmod())
```


# Exported usage

In this situation, the package is intended to provide the user with a predefined set of modules. The following example shows how to use the `.onLoad` hook function to define a module, just before sealing the namespace of the package and processing exports:

```r
# File: ./R/zzz.R

.onLoad <- function(libname, pkgname) {

  with_injector(get_default_injector(), {
    set_verbosity(0L)
    "foobar" %provides% "FOOBAR"
  })
  
  invisible(NULL)

}
```

When the package is loaded, the `"foobar"` module is defined within the scope of the default modulr injector situated outside of the package's namespace. It is then accessible by the user of the package:

```r
library(my_package)
modulr::lsmod()
```

```{r, echo = FALSE}
modulr::with_verbosity(0L, {
  modulr::define("foobar", NULL, "FOOBAR")
  modulr::lsmod()
})
```

# Modules in `inst/modules`

Finally, it is possible to define an entire set of on-disk modules at once in the `./inst/modules` directory. For instance:

```r
# File: ./inst/modules/foo.R

library(modulr)

"foo" %provides% "FOO"
```

In this case, it is necessary to add the path of the `modules` sub-directory of the installed package to the modulr's root paths. The `system.file` function is tailored for this kind of situation:

```r
# File: ./R/zzz.R

.onLoad <- function(libname, pkgname) {

  # ...

  with_injector(get_default_injector(), {
    root_config$set(c(
      root_config$get_all()[[1L]], 
      system.file("modules", package = "my_package")))
  })
  
  invisible(NULL)

}
```

Once the package is loaded, the installed `modules` directory is then reachable:

```r
library(my_package)
modulr::make("foo")
```

```{r, echo = FALSE}
modulr::reset()
modulr::define("foo", NULL, "FOO")
modulr::make("foo")
```
